---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "# ",
  fig.path = "figures/README-",
  dpi = 250
)
```


# __invgamma__

__invgamma__ implements the `(d/p/q/r)` statistics functions for the inverse gamma distribution in [R](http://cran.r-project.org).  It is ideal for using in other packages since it is lightweight and leverages the `(d/p/q/r)gamma()` line of functions maintained by CRAN.



## Getting __invgamma__

There are two ways to get __invgamma__.  For the CRAN version, use
```{r, eval=FALSE}
install.packages("invgamma")
```
For the development version, use
```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("dkahle/invgamma")
```



## The `(d/p/q/r)invgamma()` functions

The functions in __invgamma__ match those for the gamma distribution provided by the __stats__ package.  Namely, it uses as its density
_f(x) = (b^a / Gamma(a)) x^-(a+1) e^(-b/x),_
where a = `shape` and b = `rate`.



```{r}
library(invgamma)
s <- seq(0, 5, .01)
plot(s, dinvgamma(s, 7, 10), type = 'l')
```

```{r}
f <- function(x) dinvgamma(x, 7, 10)
q <- 2
integrate(f, 0, q)
(p <- pinvgamma(q, 7, 10))
qinvgamma(p, 7, 10) # = q
mean(rinvgamma(1e5, 7, 10) <= 2)
```





## The `(d/p/q/r)invchisq()` and `(d/p/q/r)invexp()` functions

















