---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "# ",
  fig.path = "figures/README-",
  dpi = 250
)
```


# __invgamma__

__invgamma__ implements the `(d/p/q/r)` statistics functions for the [inverse gamma distribution](https://en.wikipedia.org/wiki/Inverse-gamma_distribution) in [R](http://cran.r-project.org).  It is ideal for using in other packages since it is lightweight and leverages the `(d/p/q/r)gamma()` line of functions maintained by CRAN.



## Getting __invgamma__

There are two ways to get __invgamma__.  For the CRAN version, use
```{r, eval=FALSE}
install.packages("invgamma")
```
For the development version, use
```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("dkahle/invgamma")
```



## The `(d/p/q/r)invgamma()` functions

The functions in __invgamma__ match those for the gamma distribution provided by the __stats__ package.  Namely, it uses as its density
_f(x) = (b^a / Gamma(a)) x^-(a+1) e^(-b/x),_
where a = `shape` and b = `rate`.


The [PDF](https://en.wikipedia.org/wiki/Probability_density_function) (the _f(x)_ above) can be evaluated with the `dinvgamma()` function:
```{r, fig.height = 3}
library(invgamma)
library(ggplot2); theme_set(theme_bw())
x <- seq(0, 5, .01)
qplot(x, dinvgamma(x, 7, 10), geom = "line")
```


The [CDF](https://en.wikipedia.org/wiki/Cumulative_distribution_function) can be evaluated with the `pinvgamma()` function:
```{r}
f <- function(x) dinvgamma(x, 7, 10)
q <- 2
integrate(f, 0, q)
(p <- pinvgamma(q, 7, 10))
```

The [quantile function](https://en.wikipedia.org/wiki/Quantile_function) can be evaluated with `qinvgamma()`:
```{r}
qinvgamma(p, 7, 10) # = q
```

And random number generation can be performed with `rinvgamma()`:
```{r}
set.seed(1)
rinvgamma(5, 7, 10)
```
`rinvgamma()` can be used to obtain a [Monte Carlo](https://en.wikipedia.org/wiki/Monte_Carlo_method) estimate of the probability given by `pinvgamma()` above:
```{r}
samples <- rinvgamma(1e5, 7, 10)
mean(samples <= q)
```
Moreover, we can check the consistency and correctness of the implementation with
```{r, fig.height = 3}
qplot(samples, geom = "density") + 
  stat_function(fun = f,  color = "red")
```





## The `(d/p/q/r)invchisq()` and `(d/p/q/r)invexp()` functions

The [gamma distribution](https://en.wikipedia.org/wiki/Gamma_distribution) subsumes the [chi-squared](https://en.wikipedia.org/wiki/Chi-squared_distribution) and [exponential](https://en.wikipedia.org/wiki/Exponential_distribution) [distributions](https://en.wikipedia.org/wiki/Probability_distribution#Continuous_probability_distribution), so it makes sense to include the `*invchisq()` and `*invexp()` functions in __invgamma__.  Their implementations, however, wrap `*chisq()` and `*exp()`, not `*invgamma()`.  















