---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "# ",
  fig.path = "tools/README-",
  dpi = 250
)
```


# __invgamma__

__invgamma__ implements the `(d/p/q/r)` statistics functions for the [inverse gamma distribution](https://en.wikipedia.org/wiki/Inverse-gamma_distribution) in [R](http://cran.r-project.org).  It is ideal for using in other packages since it is lightweight and leverages the `(d/p/q/r)gamma()` line of functions maintained by CRAN.

_Please see the section on parameterizations below to avoid any unintended mistakes!_



### Getting __invgamma__

There are two ways to get __invgamma__.  For the [CRAN version](https://cran.r-project.org/package=invgamma), use
```{r, eval=FALSE}
install.packages("invgamma")
```
For the development version, use
```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("dkahle/invgamma")
```



### The `(d/p/q/r)invgamma()` functions

The functions in __invgamma__ match those for the gamma distribution provided by the __stats__ package.  Namely, it uses as its density
_f(x) = (b^a / Gamma(a)) x^-(a+1) e^(-b/x),_
where a = `shape` and b = `rate`.


The [PDF](https://en.wikipedia.org/wiki/Probability_density_function) (the _f(x)_ above) can be evaluated with the `dinvgamma()` function:
```{r, fig.height = 3}
library("invgamma")
x <- seq(0, 5, .01)
shape <- 7; rate <- 10
plot(x, dinvgamma(x, shape, rate), type = "l")
```


The [CDF](https://en.wikipedia.org/wiki/Cumulative_distribution_function) can be evaluated with the `pinvgamma()` function:
```{r}
f <- function(x) dinvgamma(x, shape, rate)
q <- 2
integrate(f, 0, q)
(p <- pinvgamma(q, shape, rate))
```

The [quantile function](https://en.wikipedia.org/wiki/Quantile_function) can be evaluated with `qinvgamma()`:
```{r}
qinvgamma(p, shape, rate) # = q
```

And random number generation can be performed with `rinvgamma()`:
```{r}
set.seed(1234)
rinvgamma(5, shape, rate)
```
`rinvgamma()` can be used to obtain a [Monte Carlo](https://en.wikipedia.org/wiki/Monte_Carlo_method) estimate of the probability given by `pinvgamma()` above:
```{r}
samples <- rinvgamma(1e5, shape, rate)
mean(samples <= q)
```
Moreover, we can check the consistency and correctness of the implementation with
```{r, fig.height = 3}
plot(density(samples))
curve(f(x), col = "red", add = TRUE)
```
Or
```{r}
ks.test(
  samples, 
  function(p) pinvgamma(p, shape, rate)
)
```





### The `(d/p/q/r)invchisq()` and `(d/p/q/r)invexp()` functions

The [gamma distribution](https://en.wikipedia.org/wiki/Gamma_distribution) subsumes the [chi-squared](https://en.wikipedia.org/wiki/Chi-squared_distribution) and [exponential](https://en.wikipedia.org/wiki/Exponential_distribution) [distributions](https://en.wikipedia.org/wiki/Probability_distribution#Continuous_probability_distribution), so it makes sense to include the `*invchisq()` and `*invexp()` functions in __invgamma__.  Their implementations, however, wrap `*chisq()` and `*exp()`, not `*invgamma()`.  









### A note on parameterizations

As detailed [here](https://github.com/dkahle/invgamma/issues/1), the parameterizations of the functions in this package cue off of their corresponding non-inverse distributions from **stats**. This commonly causes the confusion that, for example, the parameter `rate` in `dinvgamma()` is the rate parameter of the inverse gamma distribution. It is not! It is the rate parameter of the corresponding gamma distribution. Please take care with this distinction.


### A note on numerics

**invgamma** was intended to be a lightweight and simple, largely self-maintaining package implementing the inverse gamma, inverse chi-square, and inverse exponential distributions. It uses [the transformation theorem](https://en.wikipedia.org/wiki/Random_variable#Functions_of_random_variables) in all cases. 

One of the challenges to using naive implementations of distributions is that their [numerics](https://en.wikipedia.org/wiki/Floating-point_arithmetic) may not work well. Arithmetic on a computer is not the same as arithmetic in theory, the kind that you meet in math classes, and as a consequence the best computer implementations of mathematical facts/algorithms need to be tailored to the specific cases at hand.

In January 2023 I did a little poking around into this for `rinvgamma()` and found that it performs poorly when the shape parameter is less than .001 or so. The resulting distributions are very heavy-tailed, and the draws from these distributions returned by `rinvgamma()` are so large that they get rounded to either very large numbers (where the floating point representation of numbers does not provide many numbers) or infinity. Example:

```{r}
rinvgamma(10, shape = 3, rate = 7)
```

#### KS tests for sampling accuracy

Here is a more detailed Monte Carlo investigation that checks sampler quality using [the Kolmogorov-Smirnov test](https://en.wikipedia.org/wiki/Kolmogorov–Smirnov_test).

First, we write a basic Monte Carlo test for the sampler that works by generating a large (`n = 1e6`) sample of draws from the inverse gamma distribution for a given shape and rate:

```{r}
test_for_shape_rate <- function(shape, rate, n = 1e6) {
  samples <- rinvgamma(1e6, shape, rate)
  ks.test(samples, function(p) pinvgamma(p, shape, rate))$p.value
}

test_for_shape_rate(3, 7)
```

The function returns the $p$-value associated with the KS test, so "small" values suggest a departure from the null hypothesis that the distribution is from the corresponding inverse gamma distribution: the sampler is performing poorly. Under the null hypothesis, [the $p$-value has an approximate uniform distribution](https://en.wikipedia.org/wiki/P-value#p-value_as_the_statistic_for_performing_significance_tests), a fact that can be found in most advanced mathematical statistics books, so we would expect some proportion to be small regardless. 

We want to see the behavior of the sampler `rinvgamma()` across a wide array of parameter values. To do this, we use a range of parameter values running from small ($10^-4$) to large ($10^3$):

```{r, message=FALSE}
library("tidyverse"); library("scales")
theme_set(theme_minimal())
theme_update(panel.grid.minor = element_blank())

N <- 26
param_vals <- 10^seq(-4, 4, length.out = N)
(param_grid <- expand_grid(shape = param_vals, rate = param_vals))
```

Here's what the experiment's design space looks like:

```{r}
ggplot(param_grid, aes(shape, rate)) +
  geom_point() +
  scale_x_log10(n.breaks = 10, labels = label_number()) +
  scale_y_log10(n.breaks = 10, labels = label_number()) +
  coord_equal()
```

Now, we run our test for each point in the design space in parallel. (Note: we've suppressed warnings here that are relevant.)

```{r, warning=FALSE}
library("furrr"); furrr_options(seed = TRUE)
plan(multisession(workers = parallelly::availableCores()))

param_grid <- param_grid %>% 
  mutate(
    p_val = future_map2_dbl(shape, rate, test_for_shape_rate)
  )

plan(sequential)
```

And we visualize the distribution of the $p$-values over that space, binning the colors to at .05 to highlight the rejections of the tests at the 5% level:

```{r}
ggplot(param_grid, aes(shape, rate, color = p_val)) +
  geom_point() +
  scale_x_log10(n.breaks = 10, labels = label_number()) +
  scale_y_log10(n.breaks = 10, labels = label_number()) +
  scale_color_viridis_b(breaks = c(0, .05, 1)) +
  coord_equal()
```

If the sampler were working correctly, the $p$-values would be approximately IID uniform(0,1), so we would expect about 5% of the points to be purple, and those 5% would be uniformly distributed over the whole space with no patterns. Obviously, that's not the case: when the shape parameter is small, the test is always rejecting. Clearly, when `shape` is small, the sampler does not work well.

The cutoff seems to be right around `shape = .01`, so let's rerun the simulation zooming in on that boundary.

```{r, warning=FALSE}
# reset the grid
param_vals_small <- 10^seq(-2.5, -1.5, length.out = N)
param_grid <- expand_grid(shape = param_vals_small, rate = param_vals)


# rerun the simulation
plan(multisession(workers = parallelly::availableCores()))
param_grid <- param_grid %>% 
  mutate(
    p_val = future_map2_dbl(shape, rate, test_for_shape_rate)
  )
plan(sequential)


# plot results
ggplot(param_grid, aes(shape, rate, color = p_val)) +
  geom_point() +
  scale_x_log10(n.breaks = 10, labels = label_number()) +
  scale_y_log10(n.breaks = 10, labels = label_number()) +
  scale_color_viridis_b(breaks = c(0, .05, 1))
```


The cutoff appears to be a little below `0.01`, but I include this in the warning in `rinvgamma()` as a good rule of thumb. Let's zoom in a bit more...

```{r, warning=FALSE}
# reset the grid
param_vals_small <- 10^seq(-2.1, -2, length.out = N)
param_grid <- expand_grid(shape = param_vals_small, rate = param_vals)


# rerun the simulation
plan(multisession(workers = parallelly::availableCores()))
param_grid <- param_grid %>% 
  mutate(
    p_val = future_map2_dbl(shape, rate, test_for_shape_rate)
  )
plan(sequential)


# plot results
ggplot(param_grid, aes(shape, rate, color = p_val)) +
  geom_point() +
  scale_x_log10(n.breaks = 10, labels = label_number()) +
  scale_y_log10(n.breaks = 10, labels = label_number()) +
  scale_color_viridis_b(breaks = c(0, .05, 1))
```

The simulation confirms that over a reasonably large range of the parameters the quality of the sampler depends almost entirely on `shape` (though there is roughly a linear dependence on `shape` and `rate`). As an easy rule, the sampler can be considered unreliable for `shape` values less than `0.01`.











